---
title: "Data Visualization in R"
author: "Prof. Harbert"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    theme: united
    highlight: tango
    df_print: paged
    toc: true
    toc_float: true
    toc_depth: 4
  
---


```{r setup, include=FALSE}
library(reticulate)
use_python('/path/to/python') #EDIT
knitr::opts_chunk$set(echo = TRUE)
```



[home](https://rsh249.github.io/applied_bioinformatics)



# R Reminders

## Accessing RStudio Online

To access your RStudio account for this course navigate to http://ip.of.the.day:8787. This number can change from time to time so if you have trouble accessing this later you should email your instructor.


# Data Visualization

## Recent news:

The BBC Visual and Data Journalism team uses R and ggplot to create their publication graphics. A few weeks ago they released their toolbox on [GitHub](https://github.com/bbc/bbplot). More from [Medium](https://medium.com/bbc-visual-and-data-journalism/how-the-bbc-visual-and-data-journalism-team-works-with-graphics-in-r-ed0b35693535)



## Read Census Data

For this section we will read in a table of US Census Demographic data from https://www.kaggle.com/muonneutrino/us-census-demographic-data. The file can be found on the RStudio server at '/usr/share/data/kaggle/census.csv'. 


To read data into R we use:
read.table()
read.csv()
read.delim()

```{r, eval=FALSE}
census = read.csv('/usr/share/data/kaggle/census.csv')
head(census)

```

```{r, eval=TRUE, echo=FALSE}
census=read.csv("/Users/rharbert/Documents/census.csv")
```

LOOK BACK: How do you write a table to a new file?


## More ggplot2 Graphics

For help and an overview of all of the graphics types that are available in ggplot check out the [R ggplot2 cheatsheet](https://www.rstudio.com/wp-content/uploads/2016/11/ggplot2-cheatsheet-2.1.pdf)

```{r}
library(ggplot2)
```

### Recall: Scatterplots using 'geom_point()'

```{r}
ggplot(data = census) +
  geom_point(mapping = aes(x = Citizen, y = TotalPop))
```

## Graphs for single parameters:

### geom_histogram()

```{r}
ggplot(data=census) +
  geom_histogram(aes(x=Citizen), binwidth=100) 
```

But the x-axis is based on the maximum value, not the effective range so the distribution of the data is tough to read.

```{r}
ggplot(data=census) +
  geom_histogram(aes(x=Citizen), binwidth=100)+
  xlim(c(0,10000))
```

That also got rid of an artifact near zero. Let's inspect the data for 'Citizen' to see what data could be causing that:

```{r}
#summary(census$Citizen)
sum(census$Citizen <=0)
```

Are zero values real?

#### Sidetrack: Logical subsetting

We can use logical operators and the data.frame index notation to find ONLY those rows of data that have a zero value in the Citizen column.

```{r}
# test for Citizen ==0
logic = census$Citizen==0
# Then use this vector of logical results to index the data frame
zerocit= census[logic,] #make sure to use the rows position -> left of the column
#summary(zerocit)
```
### Density:

```{r}
ggplot(data=census) +
  geom_density(aes(x=Citizen)) +
  xlim(c(0,10000))
```

### geom_area()

```{r}
ggplot(data=census) +
  geom_area(stat='bin', bins = 30, aes(x=Citizen)) +
  xlim(c(0,10000))
```

### For Discrete values: geom_bar()

```{r}
ggplot(data=census) +
  geom_bar(aes(x=State))

```
But, this is hard to read. A few fixes:

+ Rotate x-axis labels (Use 'theme()')
+ Sort states by count (Use library(forcats) -- Tools for working with categorical variables)

```{r}
library(forcats)
ggplot(data=census) +
  geom_bar(aes(x=fct_infreq(State))) +
  xlab('') +
  theme(axis.text.x  = element_text(angle=90))
```

## Graphs for two parameters

### Continuous parameters

#### geom_point()

```{R}
ggplot(data=census) + 
    geom_point(aes(x=IncomePerCap, y=Poverty))
```

See also: geom_jitter(), and geom_label()

#### geom_smooth

```{r}
ggplot(data=census) + 
    geom_point(aes(x=IncomePerCap, y=Poverty), alpha=0.1) +
    geom_smooth(aes(x=IncomePerCap, y=Poverty)) 
```

#### Countour and density plots

```{R}
ggplot(data=census) + 
    geom_hex(aes(x=IncomePerCap, y=Poverty)) 
```
```{R}
ggplot(data=census) + 
    geom_density2d(aes(x=IncomePerCap, y=Poverty)) 
```

### One Discrete X One Continuous parameter

#### Boxplots

```{R}
ggplot(data=census[1:10000,]) + 
    geom_boxplot(aes(x=State, y=Poverty)) 
```

#### Violin charts

```{R}
ggplot(data=census[1:10000,]) + 
    geom_violin(aes(x=State, y=Poverty)) 
```

## Facetting: Multipanel plots

```{R}
ggplot(data=census[1:10000,]) + 
    geom_point(aes(x=Employed, y=Poverty)) +
    facet_grid(~State) + theme(axis.text.x = element_text(angle=90))
    
```

# Saving your workspace:

Either Session -> Save Workspace As

OR:

```{r, eval=F}
save.image('workspace_1.RData')
```

You ALSO want to save the R script file that you have been putting your code in with File-> Save(As) and using the file extension .R to identify R code files.
