---
title: "Lab 3: Advanced R"
author: "Nicholas Perry"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    theme: united
    highlight: tango
    df_print: paged
    toc: true
    toc_float: true
    toc_depth: 4
  
---


```{r setup, include=FALSE}
library(reticulate)
use_python('/path/to/python') #EDIT
knitr::opts_chunk$set(echo = TRUE)
```

#Card Game Simulation

In this lab, we will simulate how to create a deck of cards, as well as what a hand will look like.

##Create Deck

First, we need to create the deck.

```{R, eval=TRUE}
cardArray <- array(1:52)
for (i in 1:52){
  cardArray[i]=i
}
print(cardArray)
```

This code will produce a deck of 52 cards.

##Shuffle Deck

Now, we need to shuffle the deck so that all of the cards are in random places.

```{r, eval=TRUE}

for (j in cardArray){
  randomNumber <- sample(1:52, 1)
  temp = cardArray[j]
  cardArray[j]=cardArray[randomNumber]
  cardArray[randomNumber]=temp
}
print(cardArray)
```

##Deal Hand

Now, our deck is shuffled, so we can deal out a hand.

```{R, eval=TRUE}
hand <- array(1:4)
for (k in 1:5){
  hand[k]=cardArray[k]
}
print(hand)
```

##Print Hand

Now, we have our hand dealt out, so let's see what it looks like.

```{R, eval=FALSE}
for (ii in hand){
  if (((hand[ii]%%13)==0)||((hand[ii]%%13)==1)||((hand[ii]%%13)==2)||((hand[ii]%%13)==3)||((hand[ii]%%13)==4)||((hand[ii]%%13)==5)||((hand[ii]%%13)==6)||((hand[ii]%%13)==7)||((hand[ii]%%13)==8)){
    print((hand[ii]%%13)+2)
  }
  else if(hand[ii]%%13==9){
    print("Jack")
  }
  else if(hand[ii]%%13==10){
    print("Queen")
  }
  else if(hand[ii]%%13==11){
    print("King")
  }
  else if(hand[ii]%%13==12){
    print("Ace")
  }
  else{
    print("Error")
  }
}
print("of")
for (iii in hand){
  if (hand[iii]/13==0){
    print("Hearts")
  }
  else if(hand[iii]/13==1){
    print("Diamonds")
  }
  else if(hand[iii]/13==2){
    print("Spades")
  }
  else if(hand[iii]){
    print("Clubs")
  }
  else{
    print("Error")
  }
}
```

I couldn't get this to work however. I modelled everything after 13 numbers, as that's how many different cards there are. I tried using conditional statements, along with modulus and regular division, to try to demonstrate what a single card in the hand would be. However, there appears to be an error that I do not know how to solve involving the modulus math and the if statements.
